---
title: "SPARK in Staten Island"
output: pdf_document
author: Evaluated on behalf of Borough Hall and New Dorp High School by Fil Babalievsky and Atishay Sehgal
abstract: The Staten Island Borough President's Office and New Dorp High School, advised by Dr. John Ratey, have begun an evaluation of the effects of intense aerobic exercise on academic achievement.

---

# Introduction

This is a replication file for Babalievsky and Sehgal (2018), an evaluation of Project SPARK.

# Preliminaries

First we load prerequisite pack and set our working directory to the path with the data (not a local file--we cannot store the data in a public repo.)

```{r}
install.packages("tidyverse")
install.packages("stargazer")
install.packages("Rcurl")
install.packages("readxl")
library(tidyverse)
library(dplyr)
library(stargazer)
library(RCurl)
library(readxl)
setwd("~/Dropbox/New SI Stuff/SPARK Eval")
```

Next we import a function for clustered standard errors.

```{r}
# import the function from repository
# thanks https://economictheoryblog.com/2016/12/13/clustered-standard-errors-in-r/
url_robust <- "https://raw.githubusercontent.com/IsidoreBeautrelet/economictheoryblog/master/robust_summary.R"
eval(parse(text = getURL(url_robust, ssl.verifypeer = FALSE)),
     envir=.GlobalEnv)
```

# The Data

First we import data for the ninth graders in the Health Sciences SLC in New Dorp High who are enrolled in SPARK. Note that there are 57 students in SPARK and 37 students in New Dorp's 9th grade class not in SPARK. The seventh tab of the Excel file holds data for SPARK students in their first marking periods in English, math, and science. There are 171 observations, 3 classes times 51 students.

```{r}
hs_spark_1<-read_excel("SPARK 2017 HN copy.xlsx", sheet = 7, col_names = FALSE)
names(hs_spark_1)<-c("StudentID", "Course", "Mark1")
hs_spark_1
nrow(hs_spark_1)
#dummy for Spark participation
hs_spark_1$Spark<-1
#dummy for New Dorp High School students
hs_spark_1$HS<-1
```

Next we import data for ninth graders who are in the Health Sciences SLC in New Dorp High School but not in SPARK. There are 111 observations, 3 classes times 37 students.

```{r}
hs_nonspark_1<-read_excel("SPARK 2017 HN copy.xlsx", sheet = 8, col_names = FALSE)
names(hs_nonspark_1)<-c("StudentID", "Course", "Mark1")
#dummy for SPARK non-participation
hs_nonspark_1$Spark<-0
#dummy for New Dorp High School Students
hs_nonspark_1$HS<-1
nrow(hs_nonspark_1)
```


Here we combine the results from the first marking period with those of the second marking period.

```{r}
first<-rbind(hs_spark_1, hs_nonspark_1)


all_mp_2<-read_excel("SPARK MP2 Grades copy.xlsx", sheet = 1)
names(all_mp_2)
names(all_mp_2)<-c("StudentID", "Course", "Mark2")

total<-merge(first, all_mp_2, by=c("StudentID", "Course"))
total$diff<-as.numeric(as.character(total$Mark2))-as.numeric(as.character(total$Mark1))


total_hs<-total[total$HS==1,]
nrow(total_hs[total_hs$Spark==0,])
nrow(total_hs[total_hs$Spark==1,])

```


Next we look at the data from marking periods 3 and 4 and try to merge them with the data from periods 1 and 2.

```{r}
three_and_four_again<-read_excel("SPARK 2017 Term 2 MP3 and MP4 copy.xlsx", sheet = 1)
three_and_four_again
names(three_and_four_again)<-c("StudentID", "Blank1", "Blank2",  "Course", "Blank3", "Blank4", "Blank5", "Blank6", "MP3", "MP4", "Final")

total_hs<-total[total$HS==1,]
total_hs$firsfour<-substr(total_hs$Course, 1, 4)
total_hs$lastthree<-substr(total_hs$Course, 6, 8)
total_hs$Course<-paste(total_hs$firsfour, total_hs$lastthree, sep="2")
total_hs
nrow(total_hs)

Final<-left_join(total_hs, three_and_four_again, by=c("StudentID", "Course"))
Final
nrow(Final)
nrow(Final[Final$Spark==0,])
nrow(Final[Final$Spark==1,])

Final$Final <- NULL
Final

is.numeric(Final$Mark1)
is.numeric(Final$MP4)
is.numeric(Final$MP3)
is.numeric(Final$MP4)
```

Next we run our regressions. All our observations are at the student-subject level. The first regressions have the following format:

Δ=α+β×SPARK

Here Δ is the difference in grades for a student-subject between marking period 1 and a subsequent marking period, α is a constant, and β is the coefficient on a dummy variable taking value 1 if and only if the student is in SPARK. Other regressions have the format:

Δ=α+∑<sub>i</sub>(β×SPARK×Class)+∑(γ×Class)

Here, γ is the coefficient on a specific subject--Math, English, or Science. A high value of γ for, say, math, tells us that student performance declined less in

```{r}
Final$diff1<-as.numeric(as.character(Final$Mark2))-as.numeric(as.character(Final$Mark1))
Final$diff2<-as.numeric(as.character(Final$MP3))-as.numeric(as.character(Final$Mark1))
Final$diff3<-as.numeric(as.character(Final$MP4))-as.numeric(as.character(Final$Mark1))


summary(Final$diff1)

lm1<-lm(diff1~Spark,data=Final)
summary(lm1)
summary(lm1, cluster=c("StudentID"))

lm2<-lm(diff2~Spark,data=Final)
summary(lm2)

lm3<-lm(diff3~Spark,data=Final)
summary(lm3)


Final$coursechar<-as.character(Final$Course)
Final$Type<-substr(Final$coursechar, 1, 1)


Final$Spark_math<-0
Final$Spark_math[Final$Spark==1&Final$Type=="M"]<-1

Final$Spark_sci<-0
Final$Spark_sci[Final$Spark==1&Final$Type=="S"]<-1

Final$Spark_eng<-0
Final$Spark_eng[Final$Spark==1&Final$Type=="E"]<-1

Final$math<-0
Final$math[Final$Type=="M"]<-1

Final$sci<-0
Final$sci[Final$Type=="S"]<-1

Final$eng<-0
Final$eng[Final$Type=="E"]<-1


lm1_subj<-lm(diff1~Spark_math+Spark_sci+Spark_eng+math+eng,data=Final)
summary(lm1_subj)
summary(lm1_subj, cluster=c("StudentID"))

lm2_subj<-lm(diff2~Spark_math+Spark_sci+Spark_eng+math+eng,data=Final)
summary(lm2_subj)
summary(lm2_subj, cluster=c("StudentID"))

lm3_subj<-lm(diff3~Spark_math+Spark_sci+Spark_eng+math+eng,data=Final)
summary(lm3_subj)
summary(lm3_subj, cluster=c("StudentID"))
```




```{r}

Final$diff_intermediate1<-as.numeric(as.character(Final$MP3))-as.numeric(as.character(Final$Mark2))
lm_intermediate1<-lm(diff_intermediate1~Spark_math+Spark_sci+Spark_eng+math+eng,data=Final)
summary(lm_intermediate1)

Final$diff_intermediate2<-as.numeric(as.character(Final$MP4))-as.numeric(as.character(Final$MP3))
lm_intermediate2<-lm(diff_intermediate2~Spark_math+Spark_sci+Spark_eng+math+eng,data=Final)
summary(lm_intermediate2)

Final$diff_ends<-as.numeric(as.character(Final$MP4))-as.numeric(as.character(Final$Mark2))
lm_ends<-lm(diff_ends~Spark_math+Spark_sci+Spark_eng+math+eng,data=Final)
summary(lm_ends)
```

```{r}

plot(Final$Mark1)
plot(Final$Mark2)
plot(Final$MP3)
plot(Final$MP4)

```

Next we look at the kids who left our sample after Marking Period 2. Note that the SPARK students who dropped out of the sample had high MP1 grades, and an MP1 to MP2 drop that was not unusually great. We do not know why they left the sample.

```{r}
dropouts <- Final[which(is.na(Final$MP4)),]
dropouts
as.numeric(dropouts$Mark2)-as.numeric(dropouts$Mark1)
summary(as.numeric(dropouts$Mark2)-as.numeric(dropouts$Mark1))
summary(as.numeric(Final$Mark2)-as.numeric(Final$Mark1))

mean(dropouts$diff)
summary(Final$diff1)

write.csv(Final, file = "Final_Spark.csv")

```
