---
title: "SPARK Evaluation"
author: "Fil Babalievsky and Atishay Sehgal"
date: "8/2/2018"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
setwd("~/Dropbox/New SI Stuff/SPARK Eval")
Final<-read.csv("Final_SPARK.csv")
library(stargazer)
```


## Background

- Dr. John Ratey: SPARK book, compiled evidence

- Implemented in Naperville High, promising results but not randomized

- Borough Hall: Closer to a true randomized design



## Implementation

- One High School split a group of <100 ninth graders into 3

- Arbitrarily split by Gym class, no systematic differences

- Aerobic exercise (heart rate near max) rather than sports


## Empirical design

Differences in Differences: Do SPARK kids' scores grow more from Marking Period 1 to Marking Periods 2, 3, 4? Is this different in different classes?

- First: 

\[\Delta_{i,s,m}=\alpha_m+\beta_m \cdot SPARK_i\]

- Second: Split by classes

\[\Delta_{i,s,m}=\sum_j\alpha_{j,m}+\sum_j\beta_{j,m} \cdot SPARK_{i,j}\]


## Results: Simple Design

```{r, results='asis', message=FALSE, echo=FALSE}
lm1<-lm(diff1~Spark,data=Final)
lm2<-lm(diff2~Spark,data=Final)
lm3<-lm(diff3~Spark,data=Final)
stargazer(lm1, lm2, lm3, type = "latex" , header=FALSE, column.labels = c("MP1 to MP2", "MP1 to MP3", "MP1 to MP4"), dep.var.labels.include = FALSE, column.sep.width = "0.5pt", font.size="tiny")
```

## Results: Class effects

```{r, results='asis', message=FALSE, echo=FALSE}
lm1a<-lm(diff1~Spark_math+Spark_sci+Spark_eng+math+eng,data=Final)
lm2a<-lm(diff2~Spark_math+Spark_sci+Spark_eng+math+eng,data=Final)
lm3a<-lm(diff3~Spark_math+Spark_sci+Spark_eng+math+eng,data=Final)
stargazer(lm1a, lm2a, lm3a, type = "latex" , header=FALSE, column.labels = c("MP1 to MP2", "MP1 to MP3", "MP1 to MP4"), dep.var.labels.include = FALSE, column.sep.width = "0.5pt", font.size="tiny")
```

## Interpretation

- Not huge, but still interesting

- Generally larger for math, in line with prior studies


## Going forward

Ideally:

- Large scale RCT (randomly assign all SI High School gym classes to SPARK or not)

- Heterogeneous effects (measure baseline academic performance and fitness from kids, test if effects are bigger for less fit or lower scoring kids)

- Pre register the study